\bf Hierarchical Network Structure Promotes Dynamical Robustness

### 2014-11-26 tex/supplement.tex
update text

%!TEX root = ../paper.tex
\subsection{Dynamical systems on biological networks}
In the construction of a class of potential models for biological systems at any level of the biological hierarchy from metabolic to ecosystem-level networks, it is common to first attempt to define a collection of observable phenomena of interest and determine a domain (such as binary numbers, integers, or real numbers) in which each observable can be quantified. Next, it is necessary to establish the interdependencies among system components. The specific manner in which the components depend upon one another must be clarified, which is often done by determining a particular system of mathematical functions that represents a hypothesis about how the quantified observables evolve in time. To the degree to which there is uncertainty about the interactions among system components, parameters are introduced to broaden the class of models under consideration. Finally, whatever model class remains may be compared to empirical observations to determine how capable the model is of representing the phenomena of interest.

@@

A quality which is of interest to us is robustness or structural stability \cite{Smale1967}, which requires the determination of whether or not a given dynamical system that is determined to be stable remains stable under a perturbation to one or more of its defining parameters.  We will quantify this as the probability that, if some property holds for a set of parameters, it will continue to hold if we make a random perturbation about those parameter values. To do so, we will introduce, in addition to the probability distribution $\mu$ described above, a family of probability distributions $\mu'$ such that $\mu'(x,y)$ encodes the conditional probability that a system with a parameter value $x$ will have its parameter values changed to $y$ under a perturbation.  For our biological models, we will be interested in large perturbations rather than the small or infinitesimal perturbations usually considered in the theory of dynamical systems.  Then, given a binary random variable $q$, we define its robustness as the following conditional probability where $\mathbf{1}_q(x)$ is the standard indicator function equal to $1$ when $x$ satisfies $q$ and $0$ otherwise:
\begin{align}\label{eq:robustness}
  R (q,\mu') =
  \frac{\int_S d\mu(x) \int_S d\mu'(x,y) \mathbf{1}_q(x) \mathbf{1}_q(y)}
  {\int_S d\mu(x) \mathbf{1}_q(x)}
\end{align}

@@

Suppose that our Jacobian $A$ is an $n \times n$ matrix with real coefficients, connectivity $\connectivity$, and denote the proposition ``$A$ is stable'' as $\mathrm{stab}(A)$.  Then we may compute the robustness of this proposition using \ref{eq:robustness} once we have specified the probability densities $\mu$ and $\mu'$.  In general, these will depend upon the parameters of the non-linear system.  For the purpose of the current investigation, we shall make the ansatz that these are such that the distribution of entries of $A$ is (at least approximately) the uniform distribution $\mathcal{U}(-1,1)$ on the $\connectivity$-dimensional hypercube, $H^\connectivity$, of edge length $r=2$, centered about the origin. To model the perturbations, we resample the elements $k_1, k_2, \ldots k_m$ of this matrix from the same uniform distribution yielding
\begin{widetext}
$$
\mu_{k_1,\ldots,k_m}(x,y) = \prod_{y \notin \{x_1, \ldots, x_m\} } \delta(x_j-y_j) \prod_{y \notin \{x_1,\cdots,x_m\}} \mathbf{1}_{[-1,1]} (y_j) \mathbf{1}_{[-1,1]} (x_j)
$$
\end{widetext}
Under these assumptions, the expression for robustness becomes a special case of \ref{eq:robustness}:
\begin{widetext}
\begin{equation}\label{eq:condprobgen}
 R (\mathrm{stab}, \mu'_{k_1, \ldots, k_m}) =
  \frac{\int_{H^\connectivity} dx_1 \cdots dx_d \int_{H^m} {dx'}_{k_1} \cdots {dx'}_{k_m}\,
    \mathbf{1}_{stab} \mathbf{1}_{stab'}}
  {\int_{H^\connectivity} dx_1 \cdots dx_d  \, \mathbf{1}_{stab}(x_1 \cdots x_d)},
\end{equation}
\end{widetext}
where we made the abbreviations
\begin{align*}
\mathbf{1}_{stab} &= \mathbf{1}_{stab}(x_1 \cdots x_{k_1} \cdots x_{k_m} \cdots x_\connectivity), \\
\mathbf{1}_{stab'} &= \mathbf{1}_{stab'}(x_1 \cdots {x'}_{k_1} \cdots {x'}_{k_m}  \cdots x_d),
\end{align*}
i.e. $\mathbf{1}_{stab'}$ corresponds to replacing $x_{k_1}, \ldots x_{k_m}$ with their primed counterparts.

@@

This map $\hier$ is many-to-one and so there is a large class of
operations which leaves $\hier(G)$ invariant for a given graph $G$.
For instance, we may interchange the positions of the strongly
connected components relative to each other \reffighiertransformations$a$.  Leaving the components fixed, we may move links between nodes in a component \reffighiertransformations$b$ or between components, or even add or delete links \reffighiertransformations$c$.  As we shall see later, some of these operations leave invariant key dynamical quantities, so we may regard them as a symmetry groupoid of our system.

@@

The analogous results for all three variable systems are computed using Monte Carlo integration and shown in \ref{tab:structstabmat3} and \reffigrobustconnect. This process is associated with some error relative to the exact integration described above. If we sample $N = N_{\mathrm{stab}} + N_{\mathrm{unstab}}$ matrices where each has some probability $\theta$ of being stable then $N_{\mathrm{stab}}$ has a binomial distribution. We can compute a sample estimate for $\theta$, $\hat{\theta} = \frac{N_{\mathrm{stab}}}{N}$ \cite{Murphy2012}. The posterior distribution in this case is known to be a Beta distribution as a result of Beta-Binomial conjugacy
$$
\mathrm{Beta}(\theta | \mathcal{D}) = \mathrm{Beta}(\theta | N_{\mathrm{stab}} + a, N_{\mathrm{unstab}} + b)
$$
where $a$ and $b$ are the hyperparameters of the Beta prior and we consider the uninformative uniform prior corresponding to $a=b=1$. We consider the maximum a posteriori estimate
$$\hat{\theta}_{MAP} = \frac{a + N_{\mathrm{stab}} - 1}{a + b + N - 2}$$
which corresponds in this case to the maximum likelihood estimate
$$
\hat{\theta}_{MLE} = \frac{N_{\mathrm{stab}}}{N}.
$$
This estimate is characterized by the variance of the posterior Beta distribution
\begin{widetext}
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{(a+N_{\mathrm{stab}})(b+N_{\mathrm{unstab}})}{(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}})^2(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}}+1)}
$$
\end{widetext}
Since for the chosen prior $a=b=1 \ll N$ this simplifies to
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{\hat{\theta}(1-\hat{\theta})}{N}
$$
yielding the error estimate given by the associated standard deviation. In all simulations we use $N~=~10000$ so that the maximum error for $\hat{\theta}~=~0.5$ is $\sigma~=~\sqrt{\mathrm{var}(\theta | \mathcal{D})} \approx 0.005$.

>>>>

%!TEX root = ../paper.tex
\subsection{Overview}
The dynamical model given in terms of a system of differential equations for any network can be represented in terms of an interaction graph (\ref{fig:biomodelexamples} top row). These interaction graphs can be viewed as deriving from the combination of system modules that accept a given pattern of inputs and produce a given pattern of outputs (\reffigexamplesystemmodules). Symmetries are characterized by the ability to interchange these modules or their connectivity without changing some property of the system.

The network architecture can be represented in terms of an adjacency matrix and further abstracted by mapping the interaction graph to the network of strongly connected components (SCCs, see Supplementary Information) (\reffigscc). This map from the interaction graph of a biological network, referred to as $\hier$, has a collection of symmetries shown in \reffighiertransformations. These three symmetries represent transformations that can be performed on the interaction graph that do not change the network of SCCs to which it is associated \reffigscc. Two of these three are also symmetries with respect to dynamical robustness. \ref{fig:robustnesssymmetries} shows an example of these symmetries applied to a specific interaction graph.

We have derived an analytical expression for dynamical robustness, $R_{tot}$, of a biological network in terms of its interaction graph as a weighted average of the robustness of the SCCs, $R_i$, the corresponding number of links within each SCC, $d_i$, and the number of links between the SCCs, $l$. This expression is fully developed in \ref{eq:sccrobustness}, but can be schematized as in \ref{eq:robschematic} (see \reffigscc$\,$ for examples demonstrating this expression)
\begin{equation}\label{eq:robschematic}
R_{tot} = \frac{l+d_1 R_1 + d_2 R_2 + \cdots}{l+d_1 + d_2 + \cdots}.
\end{equation}
Examining this expression noting that $R_i$ are all strictly less than one proves that networks maximizing $l$, will also maximize $R_{tot}$.
This implies that the interaction graphs for systems that are the most robust will maximize the number of links between SCCs as well as the overall number of SCCs with respect to a particular system size. This analytical result predicts that any biological network whose associated dynamical system has the interaction graph \reffigscc $\,$ top will be more robust than those associated to any of the other interaction graphs in \reffigscc. Because this result is purely topological in nature, it does not depend at all upon any particular details such as the probability distribution from which the component interaction strengths are sampled or the size of the system. The result that dynamical robustness is correlated with network hierarchy therefore applies to an even broader class of dynamical systems than the particular random ensembles we have studied directly.

To test this prediction, we computed the probability distribution of stability and dynamical robustness relative to biological network architecture for ensembles of systems having two or three interacting components (see \ref{tab:structstabmat} and \ref{tab:structstabmat3}). For all of these, we found that robustness is correlated with connectivity, but that the most robust systems have intermediate connectivity for a given network size (\reffigrobustconnect). Accounting for the number of cycles in a network architecture reveals a strong correlation between robustness and connectivity that was hidden when networks with any number of cycles were considered together (\reffigconnectcycle3D3x3). While the most hierarchical network architecture will always lack cycles altogether, cycle number alone is clearly insufficient to account for robustness as the members of each class span nearly the entire range of possible robustness values. Consistent with our analysis of the symmetries of robustness, we found that the most hierarchical network architecture is the most robust (\reffigrobusthierarchy). Moreover, if we consider hierarchy partitioned by connectivity, we find that there is a monotonic increase in robustness following any line of increasing hierarchy in \reffigconnectdist3D3x3.

\subsection{Dynamical systems on biological networks}
In the construction of a class of potential models for biological systems at any level of the biological hierarchy from metabolic to ecosystem-level networks, it is common to first attempt to define a collection of observable phenomena of interest and determine a domain (such as binary numbers, integers, or real numbers) in which each observable can be quantified. Next, it is necessary to establish the interdependencies among system components. The specific manner in which the components depend upon one another must be clarified, which is often done by determining a particular system of mathematical functions that represents a hypothesis about how the quantified observables evolve in time. To the degree to which there is uncertainty about the interactions among system components, parameters are introduced to broaden the class of models under consideration. Finally, whatever model class remains may be compared to empirical observations to determine how capable the model is of representing the phenomena of interest.

@@

A quality which is of interest to us is robustness whose evaluation requires the determination of whether or not a given dynamical system that is determined to be stable remains stable under a perturbation to one or more of its defining parameters \footnote{We mean to refer to perturbations to the structure of the system itself as determined by the strengths of the couplings between the components as determined by the parameters and not only to perturbations of the state vector at a given point in time. For example, when the network of components and interactions corresponds to a gene-regulatory network, then mutation is one mechanism by which such perturbations may arise}.  We will quantify this as the probability that, if some property holds for a set of parameters, it will continue to hold if we make a random perturbation about those parameter values. To do so, we will introduce, in addition to the probability distribution $\mu$ described above, a family of probability distributions $\mu'$ such that $\mu'(x,y)$ encodes the conditional probability that a system with a parameter value $x$ will have its parameter values changed to $y$ under a perturbation.  For our biological models, we will be interested in large perturbations rather than the small or infinitesimal perturbations usually considered in the theory of dynamical systems.  Then, given a binary random variable $q$, we define its robustness as the following conditional probability where $\mathbf{1}_q(x)$ is the standard indicator function equal to $1$ when $x$ satisfies $q$ and $0$ otherwise:
\begin{align}\label{eq:robustness}
  R (q,\mu') =
  \frac{\int_S d\mu(x) \int_S d\mu'(x,y) \mathbf{1}_q(x) \mathbf{1}_q(y)}
  {\int_S d\mu(x) \mathbf{1}_q(x)}
\end{align}

@@

Suppose that our Jacobian $A$ is an $n \times n$ matrix with real coefficients, connectivity $\connectivity$, and denote the proposition ``$A$ is stable'' as $\mathrm{stab}(A)$.  Then we may compute the robustness of this proposition using \ref{eq:robustness} once we have specified the probability densities $\mu$ and $\mu'$.  In general, these will depend upon the parameters of the non-linear system.  For the purpose of the current investigation, we shall make the ansatz that these are such that the distribution of entries of $A$ is (at least approximately) the uniform distribution $\mathcal{U}(-1,1)$ on the $\connectivity$-dimensional hypercube, $H^\connectivity$, of edge length $r=2$, centered about the origin. To model the perturbations, we resample the elements $k_1, k_2, \ldots k_m$ of this matrix from the same uniform distribution yielding
\begin{widetext}
$$
\mu_{k_1,\ldots,k_m}(x,y) = \prod_{y \notin \{x_1, \ldots, x_m\} } \delta(x_j-y_j) \prod_{y \notin \{x_1,\cdots,x_m\}} \mathbf{1}_{[-1,1]} (y_j) \mathbf{1}_{[-1,1]} (x_j)
$$
\end{widetext}
Under these assumptions, the expression for robustness becomes a special case of \ref{eq:robustness}:
\begin{widetext}
\begin{equation}\label{eq:condprobgen}
 R (\mathrm{stab}, \mu'_{k_1, \ldots, k_m}) =
  \frac{\int_{H^\connectivity} dx_1 \cdots dx_d \int_{H^m} {dx'}_{k_1} \cdots {dx'}_{k_m}\,
    \mathbf{1}_{stab} \mathbf{1}_{stab'}}
  {\int_{H^\connectivity} dx_1 \cdots dx_d  \, \mathbf{1}_{stab}(x_1 \cdots x_d)},
\end{equation}
\end{widetext}
where we use the abbreviations
\begin{align*}
\mathbf{1}_{stab} &= \mathbf{1}_{stab}(x_1 \cdots x_{k_1} \cdots x_{k_m} \cdots x_\connectivity), \\
\mathbf{1}_{stab'} &= \mathbf{1}_{stab'}(x_1 \cdots {x'}_{k_1} \cdots {x'}_{k_m}  \cdots x_d),
\end{align*}
i.e. $\mathbf{1}_{stab'}$ corresponds to replacing $x_{k_1}, \ldots x_{k_m}$ with their primed counterparts.

@@

This map $\hier$ is many-to-one and so there is a large class of
operations which leaves $\hier(G)$ invariant for a given graph $G$.
For instance, we may interchange the positions of the strongly
connected components relative to each other \reffighiertransformations$a$.  Leaving the components fixed, we may move links between nodes in a component \reffighiertransformations$b$ or between components, or even add or delete links \reffighiertransformations$c$.  As we shall see later, some of these operations leave invariant key dynamical quantities, in our case dynamical robustness, so we may regard them as a symmetry groupoid of our system with respect to the robustness property.

@@

The analogous results for all three variable systems are computed using Monte Carlo integration and shown in \ref{tab:structstabmat3} and \reffigrobustconnect. This process is associated with some error relative to the exact integration described above. In all simulations we use $N~=~10000$ so that the maximum error for $\hat{\theta}~=~0.5$ is $\sigma~=~\sqrt{\mathrm{var}(\theta | \mathcal{D})} \approx 0.005$ (see \emph{Methods}).
